---
title: "R Notebook"
output: html_notebook
---
# Biostatistics Project - Seed dispersal data analyses

Clears Environment
```{r}
rm(list = ls())
```

```{r}
#install.packages(c("ggraph", "tidygraph", "igraph"))

library(igraph)
library(tidygraph)
library(ggraph)
```


### imports and merges the data for Jamaica
```{r}
# Set the folder path
folder_path <- "web-of-life_2026-03-18_071546"
file_names <- c("M_SD_003.csv", "M_SD_004.csv", "M_SD_005.csv", "M_SD_006.csv")

# Read all files and store them in a list
all_networks <- list()
for(file in file_names) {
  name <- gsub(".csv", "", file)
  all_networks[[name]] <- read.csv(file.path(folder_path, file))
  
  # Print dimensions to check
  cat(name, "dimensions:", dim(all_networks[[name]]), "\n")
}

# Method 1: Using dplyr and tidyr (recommended)
library(dplyr)
library(tidyr)

# Convert each network to long format and combine
combined_long <- data.frame()

for(network_name in names(all_networks)) {
  df <- all_networks[[network_name]]
  
  # Get plant names from first column (assuming it's named 'X' or similar)
  plant_col <- names(df)[1]  # First column name (likely 'X')
  
  # Convert to long format
  df_long <- df %>%
    pivot_longer(cols = -all_of(plant_col),  # All columns except plant column
                 names_to = "bird_species",
                 values_to = "interaction") %>%
    rename(plant_species = all_of(plant_col)) %>%
    mutate(network = network_name)
  
  # Combine with main dataframe
  combined_long <- bind_rows(combined_long, df_long)
}

# Sum interactions by plant and bird species
merged_data <- combined_long %>%
  group_by(plant_species, bird_species) %>%
  summarise(interaction = sum(interaction, na.rm = TRUE), .groups = "drop")

# Convert back to wide format (plants as rows, birds as columns)
merged_wide <- merged_data %>%
  pivot_wider(names_from = bird_species,
              values_from = interaction,
              values_fill = NA)  # Use NA for missing combinations

# View the merged data
print("Merged data dimensions:")
dim(merged_wide)
print(head(merged_wide, 10))


```


## Bipartite network of plant-bird interaction 

Since the data is for only two trophic levels i.e. plants and frugivore birds, the bipartite network was made. 
```{r}
library(bipartite)
set.seed(123)

merged_wide <- as.data.frame(merged_wide)
rownames(merged_wide) <- merged_wide[, 1]
merged_wide <- merged_wide[, -1]

merged_wide[is.na(merged_wide)] <- 0

```
```{r}

plotweb(merged_wide , horizontal = TRUE, spacing = 0.9, text_size = 0.5, sorting = "increasing", scaling = "relative", curved_links = TRUE, higher_color = "maroon", lower_color = "maroon", link_color = "pink", link_alpha = 0.6, lower_border = "black", higher_border = "black")


```
**Figure 1:** Bipartite Network


##Basic Stiatistics
```{r}
n_plants <- dim(merged_wide)[1]  #number of plant species
print("number of plant species")
n_plants
n_birds <- dim(merged_wide)[2] #number of birds species
print("number of birds species")
n_birds
s <- n_plants + n_birds #total number of species
print("total number of species")
s
n_inter <- sum(merged_wide) #total number of interactions
print("total number of interactions")
n_inter
```
### Connectance

```{r}
C <- networklevel(merged_wide, index = "connectance")
print("Connectance")
C
```
Tells how dense the network is. 

### Nestedness

```{r}
NODF <- networklevel(merged_wide, index = "nestedness")
print("Nestedness (NODF)")
NODF
```
indicates:
- generalist–specialist structure
- stability of network

### Specialization(H2')
```{r}
H2 <- H2fun(merged_wide)
print("Network specialization (H2')")
H2
```


### Interaction evenness

```{r}
IE <- networklevel(merged_wide, index = "interaction evenness")
print("Interaction evenness")
IE
```

tells if interactions are evenly distributed or dominated. 

### Modularity

```{r}
mod <- computeModules(merged_wide)
print("Modularity")
mod@likelihood

plotModuleWeb(mod)
```
**Figure 2:** one of the example module is shown here

### Modules 
```{r}
cz <- czvalues(mod)
head(cz)
modules <- mod@modules

# recreate web from your merged data
web <- as.matrix(merged_wide)

mode(web) <- "numeric"
web[is.na(web)] <- 0

# remove empty species
web <- web[rowSums(web) > 0, colSums(web) > 0]
# convert to graph
edges <- as.data.frame(web) |>
  tibble::rownames_to_column("Plant") |>
  tidyr::pivot_longer(-Plant, names_to = "Bird", values_to = "Count") |>
  dplyr::filter(Count > 0)

g <- graph_from_data_frame(edges, directed = FALSE)

# detect communities
comm <- cluster_fast_greedy(g)

membership <- membership(comm)
module_df <- data.frame(
  species = names(membership),
  module  = membership
)

head(module_df)
split(module_df$species, module_df$module)

length(unique(membership))
```
From here we see that the species interactions can be divided into 7 modules.

```{r}
set.seed(123)
edges <- as.data.frame(web) %>%
  tibble::rownames_to_column("Plant") %>%
  pivot_longer(-Plant, names_to = "Bird", values_to = "Count") %>%
  filter(Count > 0)
g <- graph_from_data_frame(edges, directed = FALSE)

# assign weights
E(g)$weight <- edges$Count
comm <- cluster_fast_greedy(g)
membership <- membership(comm)
tg <- as_tbl_graph(g) %>%
  mutate(
    deg = centrality_degree(),
    membership = membership[name]   # attach module info
  )

ggraph(tg, layout = "fr") +
  geom_edge_link(alpha = 0.2, colour = "black") +
  geom_node_point(aes(size = deg, color = as.factor(membership))) +
  scale_size(range = c(2, 8)) +
  scale_color_brewer(palette = "Dark2") +
  theme_void()
```
**Figure 3:** To visualize the 7 modules.



```{r}
# assign node type (plants vs birds)
V(g)$type <- V(g)$name %in% rownames(web)

# ensure edge weights exist
E(g)$weight <- E(g)$weight %||% 1   # fallback if missing

tg <- as_tbl_graph(g) %>%
  mutate(
    type_label = ifelse(type, "Plant", "Bird"),
    deg = centrality_degree()
  )
ggraph(tg, layout = "fr") +
  geom_edge_link(aes(width = weight),
                 alpha = 0.3,
                 colour = "black") +
  
  geom_node_point(aes(size = deg,
                      color = type_label)) +
  
  scale_color_manual(values = c(
    "Plant" = "forestgreen",
    "Bird"  = "steelblue"
  )) +
  
  scale_size(range = c(2, 8)) +
  scale_edge_width(range = c(0.2, 2)) +
  
  theme_void() +
  theme(legend.position = "right")
```
**Figure 4:**here the birds and the plants are depicted differently and from the size of the nodes we can say that few birds are more generalized while most plants show higher degree i.e. more generality. Higher degree means more resilient the system is. 

```{r}
edges <- as.data.frame(web) %>%
  tibble::rownames_to_column("Plant") %>%
  tidyr::pivot_longer(-Plant, names_to = "Bird", values_to = "Count") %>%
  dplyr::filter(Count > 0)

g <- graph_from_data_frame(edges, directed = FALSE)

# assign weights properly
E(g)$weight <- edges$Count

tg <- as_tbl_graph(g) %>%
  mutate(
    type_label = ifelse(name %in% rownames(web), "Plant", "Bird"),
    deg = centrality_degree()
  )

ggraph(tg, layout = "fr") +
  geom_edge_link(aes(width = weight),
                 alpha = 0.4,
                 colour = "black") +
  
  geom_node_point(aes(color = type_label),
                  size = 4) +   # fixed size as you wanted
  
  scale_color_manual(values = c(
    "Plant" = "forestgreen",
    "Bird"  = "steelblue"
  )) +
  
  scale_edge_width(range = c(0.2, 3)) +   # wider range for visibility
  
  theme_void() +
  theme(legend.position = "right")
```
**Figure 5:** here we see that the more connected species are at the center while the less conected, more speciazed ones are to the outside. 


## Network level analyses
```{r}

Network_analyses <- networklevel(merged_wide)
Network_analyses <- data.frame(Network_analyses)
```

The network exhibited moderate connectance (0.115) and nestedness (NODF = 38.09), indicating a structured yet flexible interaction pattern. Modularity was relatively high (Q = 0.45), suggesting the presence of distinct interaction compartments. Specialization (H2′ = 0.40) was moderate, reflecting a balance between generalist and specialist species. High robustness values (0.64–0.73) indicate resilience to species loss, consistent with the observed nested organization.




## Species level metrics
```{r}
web <- web[rowSums(web) > 0, colSums(web) > 0]
dim(web)

sp <- specieslevel(web)
sp
```


```{r}
plants_df <- as.data.frame(sp$`higher level`)
plants_df$Species <- rownames(plants_df)

birds_df <- as.data.frame(sp$`lower level`)
birds_df$Species <- rownames(birds_df)

write.csv(plants_df, "plant_species_metrics.csv", row.names = FALSE)
write.csv(birds_df, "bird_species_metrics.csv", row.names = FALSE)
```
Species-level analysis revealed a heterogeneous interaction structure, with a few highly connected generalist species and many low-connectivity specialists. Species strength indicated the presence of key plant and bird species that likely function as keystone nodes in the network. Specialization values (d′) suggested a mixture of generalist and specialist strategies, consistent with the observed moderate network-level specialization (H2′). This structure supports a nested organization, enhancing network robustness and resilience to species loss.



```{r}
plants <- data.frame(
  Species  = rownames(sp$`higher level`),
  Strength = sp$`higher level`$species.strength,
  Degree   = sp$`higher level`$degree,
  d        = sp$`higher level`$d
)

birds <- data.frame(
  Species  = rownames(sp$`lower level`),
  Strength = sp$`lower level`$species.strength,
  Degree   = sp$`lower level`$degree,
  d        = sp$`lower level`$d
)

# normalize function
norm <- function(x) (x - min(x)) / (max(x) - min(x))

plants$keystone_score <- norm(plants$Strength) +
                         norm(plants$Degree) -
                         norm(plants$d)

birds$keystone_score  <- norm(birds$Strength) +
                         norm(birds$Degree) -
                         norm(birds$d)

# top 10 plants
top_plants <- plants[order(-plants$keystone_score), ][1:10, ]

# top 10 birds
top_birds  <- birds[order(-birds$keystone_score), ][1:10, ]

top_plants
top_birds
```























